Density-dependent selection in evolutionary genetics: a lottery model of Grime's triangle

### 2016-10-27 manuscript.tex
latest

Nevertheless, there are strong indications there are broader principles governing the operation of selection. In many groups of organisms (including corals \citep{darling_2012}, insects \citep{southwood_1977}, fishes \citep{winemiller_1992} and plants \citep{grime_1988}), species can be partitioned into a small number of distinct trait clusters corresponding to fundamentally distinct ``primary strategies'' \citep{winemiller_2015}. The most famous example is Grime's plant trait classification scheme \citep{grime_1974,grime_1977,grime_1988}. Grime considered two broad determinants of population density: stress (persistent hardship e.g. due to resource scarcity, unfavorable temperatures or toxins) and disturbance (intermittent destruction of vegetation e.g. due to trampling, herbivory, pathogens, extreme weather or fire).  The extremes of these two factors define three primary strategies denoted by C/S/R respectively: competitors ``C'' excel in low stress, low disturbance environments; stress tolerators ``S'' excel in high stress, low disturbance environments; and ruderals ``R''  excel in low stress, high disturbance environments. Survival is not possible in high-stress, high-disturbance environments. Grime showed that measures of C, S and R across a wide range of plant species are anti-correlated, so that strong C-strategists are weak S and R strategists, and so on. Thus, plant species can be classified on a triangular C/S/R ternary plot \citep{grime_1974}. Trait classification schemes for other organisms closely parallel Grime's scheme \citep{winemiller_2015}.

@@

\section*{The Model}

@@

\subsection*{Environmental archetypes and Grime's triangle}

@@

The stress archetype is more ambiguous, and has been the subject of an extensive debate in the plant ecology literature (the ``Grime-Tilman'' debate \citep{aerts_1999}). In Grime's view, stressful environments impose such severe challenges that surviving the stressors at all is the primary challenge. In our model, this can be expressed by $b_i\ll 1$ and $b_i/d_i\approx 1$. The propagule density $L$ (as well as $N/K$) are suppressed to such low levels that there is essentially no competition between individuals. The severity of the stress is such that mutations which appreciably improve $b_i$ are extremely rare, so $b_i$ is constrained to remain low.

The alternative view is that the stress archetype should rather be interpreted as a large reduction in the maximum number of individuals that can be supported by the environment \citep{taylor_1990}. For example, in the commonly cited case that the stress is induced by a scarcity of consumable resources, competition for resources would likely be intense, and the stressed population should actually be regarded as having a high population density. In our model, this would imply a large reduction in $K$ (greater per-individual territorial requirement). That is, $N$ under stress is much lower than  under ideal conditions, but it is not much lower than $K$ for the stressful environment. Since our model accommodates both of these alternatives, we include them both here.

The mapping of environmental achetypes to our model parameters is summarized in the first two rows of Fig. \ref{fig:table}. Also shown is the approximate dependence of $\Delta_+ n_i$ on $b_i$ and $c_i$ for each archetype (third row), which can be used infer the expected direction of evolution for the traits $b$, $c$ and $d$ (fourth row).

The latter is obtained as follows. As noted in the previous section, if beneficial mutations can survive the low-abundance stochastic regime, their behavior is governed deterministically by Eq. \eqref{eq:master}. They will then proceed to grow deterministically (establishment). The probability of establishment increases with the mutant fitness advantage, and is therefore typically on the order of a few percent, whereas the fixation of neutral mutations is exceedingly unlikely (probability of order $1/N$). Consequently, the direction of evolutionary change is determined by which trait changes confer an appreciable benefit, subject to the constraints imposed by the environment.

For example, in Grime's version of the stress archetype, population density is low, so competition is not important, and so only mutants with greater $b$ or lower $d$ will have an appreciably greater $\Delta n_i$. Mutations in $c$ are effectively neutral, and will rarely fix. However, by definition of the stress archetype, $b$ is constrained to be very small. Thus, while some rare mutations may produce small improvements in $b$, it is much more likely that mutations will arise that lower $d$, making this the expected direction of evolutionary change for Grime's stress archetype.

\begin{figure}
\centering
\begin{tabular}{*{5}{c}}
  & Ideal & Disturbance* & Stress (G) & Stress (K) \\ \hline
  Parameter- & $d_i \ll 1$ & $d_i \approx 1$ & $b_i \ll 1$ & $b_i \ll 1$ \\
  regime & $b_i/d_i\gg 1$ & $b_i/d_i\gg 1$ & $b_i\approx d_i$ & $b_i>  d_i$ \\
  Density $N/K$  & High & Low & Low & High \\
  $\Delta_+ n_i\propto$ & $b_i c_i$ & $b_i$ & $b_i$ & $b_i c_i$ \\
  Evolution for & $\uparrow b$, $ \uparrow c$ & $\uparrow b$, $\downarrow d$ & $\downarrow d$ & $\uparrow c$
\end{tabular}
\caption{\label{fig:table} The realization of Grime's environmental archetypes in our model, as well as the low-$K$ variant of the stress archetype. Shown are the mapping to our parameters of each archetype, the approximate dependence of $\Delta_+ n_i$ on $b_i$ and $c_i$, as well as the corresponding expected evolutionary changes in $b_i$, $c_i$ and $d_i$. *Mortality affects both adults and juveniles in the disturbance archetype, with $\Delta_+ n_i$ replaced by $(1-d_i)\Delta_+ n_i$ in Eq. \eqref{eq:delttot}.}
\end{figure}

@@

How does Fig. 4 compare to Grime's C/S/R strategies? In disturbed environments, the ruderal (R) strategy is characterized by high fecundity, wide dispersal and short generation times.

We predict evolution for higher $b$, but not higher $c$ or lower $d$ (constrained to be large), amounts to focusing on fecundity. While there is no dispersal limitation in our model, propagules must be minimally effective at dispersal since our model only includes propagules which will enter territorial contests on unoccupied sites;

 Shorter generation times are thought to be a mechanism for increasing the chance that the reproductive cycle completes before disturbance-induced death

The evolution of the generation time itself is beyond the scope of our model, but it could

>>>>

Nevertheless, there are strong indications there are broader principles governing the operation of selection. In many groups of organisms, including corals \citep{darling_2012}, insects \citep{southwood_1977}, fishes \citep{winemiller_1992}, zooplankton \citep{allan_76} and plants \citep{grime_1988}, different species can be divided into a small number of distinct trait clusters corresponding to fundamentally distinct ``primary strategies'' \citep{winemiller_2015}. The most famous example is Grime's plant trait classification scheme \citep{grime_1974,grime_1977,grime_1988}. Grime considered two broad determinants of population density: stress (persistent hardship e.g. due to resource scarcity, unfavorable temperatures or toxins) and disturbance (intermittent destruction of vegetation e.g. due to trampling, herbivory, pathogens, extreme weather or fire).  The extremes of these two factors define three primary strategies denoted by C/S/R respectively: competitors ``C'' excel in low stress, low disturbance environments; stress tolerators ``S'' excel in high stress, low disturbance environments; and ruderals ``R''  excel in low stress, high disturbance environments. Survival is not possible in high-stress, high-disturbance environments. Grime showed that measures of C, S and R across a wide range of plant species are anti-correlated, so that strong C-strategists are weak S and R strategists, and so on. Thus, plant species can be classified on a triangular C/S/R ternary plot \citep{grime_1974}. Trait classification schemes for other organisms closely parallel Grime's scheme \citep{winemiller_2015}.

@@

\section*{Model}\label{sec:model}

@@

\subsection*{Grime's triangle}\label{sec:grime}

@@

The stress archetype is more ambiguous, and has been the subject of an extensive debate in the plant ecology literature (the ``Grime-Tilman'' debate \citep{aerts_1999}). In Grime's view, stressful environments impose such severe challenges that they are only barely survivable.

In our model, this can be expressed by $b_i\ll 1$ and $b_i/d_i\approx 1$. The propagule density $L$ (as well as $N/K$) are suppressed to such low levels that there is essentially no competition between individuals. The severity of the stress is such that mutations which appreciably improve $b_i$ are extremely rare, so $b_i$ is constrained to remain low.

The alternative view is that the stress archetype should rather be interpreted as a reduction in the maximum number of individuals that can be supported by the environment \citep{taylor_1990}; in our model, this means a greater per-individual territorial requirement represented by a lower $K$ for a given amount of space. The stress implies that $b$ is low

For example, in the commonly cited case that the stress is induced by a scarcity of consumable resources, it is argued that competition for resources would likely be intense, and the stressed population should actually be regarded as having a high population density \citep{taylor_1990}.

That is, $N$ under stress is much lower than  under ideal conditions, but it is not much lower than $K$ for the stressful environment. Since our model accommodates both of these alternatives, we include them both here.

The mapping of environmental achetypes to our model parameters is summarized in the first two rows of Fig. \ref{fig:table}. Also shown is the approximate dependence of $\Delta_+ n_i$ on $b_i$ and $c_i$ for each archetype (third row). These can be used infer the expected direction of evolution for the traits $b$, $c$ and $d$ (fourth row) as follows.

As noted in the previous section, if beneficial mutations establish (i.e survive the low-abundance stochastic regime), they will proceed to grow deterministically according to Eq. \eqref{eq:master}. The probability of establishment increases with the mutant fitness advantage, and is therefore typically on the order of one percent, whereas the fixation of neutral mutations is exceedingly unlikely (probability of order $1/N$). Consequently, the direction of evolutionary change is determined by which trait changes confer an appreciable benefit, subject to the constraints imposed by the environment.

For example, in Grime's version of the stress archetype, $L$ is low, so competition is not important, and only mutants with greater $b$ or lower $d$ will have an appreciably greater $\Delta n_i$. Mutations in $c$ are effectively neutral, and will rarely fix. However, by definition of the stress archetype, $b$ is constrained to be small. Thus, while some rare mutations may produce small improvements in $b$, it is much more likely that mutations will arise that lower $d$, making this the expected direction of evolutionary change for Grime's stress archetype.

\begin{figure}
\centering
\begin{tabular}{*{5}{c}}
  & Ideal & Disturbance* & Stress (G) & Stress (K) \\ \hline
  Parameter- & $d \ll 1$ & $d \approx 1$ & $b \ll 1$ & $b \ll 1$ \\
  regime & $b/d\gg 1$ & $b/d\gg 1$ & $b\approx d$ & $b>d$ \\
  Density $N/K$  & High & Low & Low & High \\
  $\Delta_+ n_i\propto$ & $b_i c_i$ & $b_i$ & $b_i$ & $b_i c_i$ \\
  Evolution for & $\uparrow b$, $ \uparrow c$ & $\uparrow b$, $\downarrow d$ & $\downarrow d$ & $\uparrow c$
\end{tabular}
\caption{\label{fig:table} The realization of Grime's environmental archetypes in our model, as well as the low-$K$ variant of the stress archetype. Shown are the mapping to our parameters of each archetype, the approximate dependence of $\Delta_+ n_i$ on $b_i$ and $c_i$, as well as the corresponding expected evolutionary changes in $b_i$, $c_i$ and $d_i$. *Mortality affects both adults and juveniles in the disturbance archetype, with $\Delta_+ n_i$ replaced by $(1-d_i)\Delta_+ n_i$ in Eq. \eqref{eq:delttot}.}
\end{figure}

@@

How does Fig. 4 compare to Grime's C/S/R strategies? In disturbed environments, we predict evolution for higher $b$ and lower $d$, but not higher $c$. Higher $b$ means higher fecundity, but not necessarily mass propagule production: $b$ represents only those propagules which sucessfully develop into juveniles in unoccupied territories. This is broadly consistent with the ruderal primary strategy. Plant ruderals devote a large proportion of their productivity to seed production \cite{grime_1977}, whereas the analogous ``opportunistic'' strategists in fishes have large intrinsic growth rates \citep{winemiller_1992}. In corals, a distinguishing feature of the ruderal cluster is brood spawning (rather than broadcast spawning), which corresponds to higher parental investment and lower overall propagule production, but may be important for successful propagation in low density, disturbed environments due to Allee effects \citep{darling_2012}. Lower $d$ could be achieved by improved individual resistance to physical destruction, but it is hard to reduce mortality in the face of severe disturbances. Given this constraint, shortening the time to reproductive maturity (the iteration time in our model) is an effective way of reducing the chance of death per iteration $d$. A short life cycle is probably the most consistent characteristic of ruderals \cite{grime_1977,winemiller_1992,darling_2012}.

In stressful environments, we predict evolution either for lower $d$ or higher $c$ depending on how the stress interpreted.

In ideal environments, we predict evolution for higher $b$ and $c$, but not lower $d$ (which is already small). In plants and corals, the ability to secure territories $c$ is strongly associated with rapid individual growth, and rapid individual growth is a defining trait of the competitor trait cluster \cite{grime_1977,winemiller_1992,darling_2012}.
